---
title: "ENVS-193DS_Infographic_Visualizations"
author: "Ella Menin"
date: "2026-05-28"
format:
  pdf:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show up
---

# Setting up

```{r}
#| label: reading-in-packages

# reading in package: ggridges
library(ggridges)
# reading in package: dplyr
library(dplyr)
# reading in package: janitor
library(janitor)
# reading in package: ggplot2
library(ggplot2)
```

```{r}
#| label: reading-in-personal-data

# creating object personal_data from reading in personal data file
personal_data <- read.csv("data/personal_data.csv")

```

```{r}
#| label: cleaning-data

# creating new object my_data_clean from personal_data data frame
my_data_clean <- personal_data |> 
  # using janitor to clean names (underscores and all lowercase)
  janitor::clean_names() |> 
  # using filter removing certain rows 
  filter(!is.na(temperature_h_w_c), # removes rows where temperature is n/a
         temperature_h_w_c != "") |> # removes rows where temperature is an empty string
  # mutating temperature variable to change levels, influences the order in which they will appear in a plot
  mutate(temperature_h_w_c = factor(temperature_h_w_c,
                                    # levels go hot -> warm -> cold
                                    levels = c("H", "W", "C"),
                                    # changing from letters to whole words (easier to understand in plot)
                                    labels = c("Hot", "Warm", "Cold")))
  
```

# Creating visualizations!

## Visualization 1: Jitter plot -- relationship between water intake and temperature that day

```{r}

# base layer ggplot with my_data_clean data frame
ggplot(data = my_data_clean,
       # x axis is temperature (hot, warm, or cold)
      mapping = aes(x = temperature_h_w_c,
           # y axis is water intake in liters
           y = water_intake_l,
           # color by temperature
           color = temperature_h_w_c)) +
  # first layer jitter to show individual data points
  geom_jitter(width = 0.15, # controlling horizontal spread
              height = 0, # controlling vertical spread
              size = 3, # making the dots easy to see
              alpha = 0.7) + # slight transparency of dots
  # adding in the mean point to make it easier to read
  stat_summary(fun = mean,
               # point to represent mean
               geom = "point",
               # mean point size bigger than individual point size
               size = 5,
               # making mean points black so they can easily be distinguished
               color = "black") +
  # manually setting colors for temperature
  scale_color_manual(values = c(
    # hot is a red color because duh
    "Hot" = "tomato2",
    # warm is a golden color because it is warm but not too... hot
    "Warm" = "goldenrod2",
    # cold is a blue color because duh
    "Cold" = "steelblue3"
  )) +
  # setting title, subtitle, and axes titles
  labs(
    # title described the two variables being related
    title = "Water intake varied with temperature experience",
    # decided to throw in a subtitle talking about what is seen in the plot
    subtitle = "Hotter days were generally associated with greater hydration",
    # x axis is temperature experience 
    x = "Temperature experience",
    # y axis is water intake in liters
    y = "Water intake (L)"
  ) +
  # cute little theme 
  theme_classic(base_size = 14) + # base size 14 so components can easily be read
  # removing legend because i do not find it to be necessary here
  theme(legend.position = "none")

```


## Visualization 2: Bubble plot -- daily steps, water intake, shift duration, and illness

```{r}

# creating new object bubble data out of my_data_clean data frame for bubble plot
bubble_data <- my_data_clean |>
  # using filter function to get tell r to ignore n/a values for following variables so that if anything is missing there will be no issue
  filter(
    !is.na(steps), # ignore n/a for steps
    !is.na(water_intake_l), # ignore n/a for water intake
    !is.na(duration_of_shift_hrs), # ignore n/a for duration of shift
    !is.na(sick_y_n) # ignore n/a for sick
  )

```


```{r}

# further editing bubble_data data frame because first draft of graph looked messy
bubble_data <- bubble_data |>
  # using mutate function to 
   mutate(
     # removes commas from steps column and converts to numercial format rather than character 
    steps = as.numeric(gsub(",", "", steps)),
    # creating new variable where steps are /1000 to make x axis cleaner in plot
    steps_k = steps / 1000
  )
  
# base layer ggplot with data frame bubble_data
ggplot(data = bubble_data,
       # aes to set axes
       mapping = aes(x = steps, # x axis is daily steps
           y = water_intake_l, # y axis is water intake in liters
           size = duration_of_shift_hrs, # size of bubbles changes by the duration of my shift in hours
           fill = sick_y_n)) + # color is filled by whether or not i am sick
  # first layer geom_point to show individual data 
  geom_point(
    # shape is style 21, which is a filled circle with a colored border, or in this case...
    shape = 21,
    # ...border is white so it looks cool when they overlap!
    color = "white",
    # make the points see through so you can see if there are any overlapping
    alpha = 0.75,
    # controls thickness of border around data points
    stroke = 1
  ) +
  # custom colors
  scale_fill_manual(values = c(
    # yes i have a sinus infection = red
    "Y" = "tomato2",
    # no i do not have a sinus infection = blue
    "N" = "steelblue3"
  )) +
  # customizing bubble size scale
  scale_size_continuous(
    # controls minimum and maximum bubble sizes so smaller values result in smaller circles
    range = c(4, 12),
    # specifies which shift duration values should appear in bubble size legend
    breaks = c(2, 4, 6, 8)
  ) +
  # setting titles
  labs(
    # title states a general idea taken from the plot
    title = "More demanding days required more hydration",
    # subtitle informs reader what bubble size represents more clearly
    subtitle = "Bubble size represents work shift duration",
    # x axis is daily steps measured in thousands of steps
    x = "Daily steps (thousands)",
    # y axis is water intake measured in liters
    y = "Water intake (L)",
    # label to explain that color based on whether or not i had a sinus infection :P
    fill = "Sinus infection",
    # label to explain that size of bubble is based on shift duration with legend
    size = "Shift duration (hrs)"
  ) +
  # nice theme with base size 14 for easy readibility
  theme_classic(base_size = 14) +
  # had to add this in because my legend bubbles kept DISAPPEARING 
  guides(
    # shape 21 for legend bubbles as well
  size = guide_legend(override.aes = list(shape = 21,
                                          fill = "grey70")) # fill with grey
)

```


## Visualization 3: Ridgeline plot -- showing how much water I drink when I have a sinus infection vs do not have a sinus infection

```{r}

# base layer ggplot using my_data_clean data frame
ggplot(data = my_data_clean,
       # setting x and y axes
       mapping = aes(x = water_intake_l, # x axis water intake in liters
           y = sick_y_n, # y axis sick yes or no 
           fill = sick_y_n)) + # fill color by sick yes or no
  # first layer setting ridgeline densities
  geom_density_ridges(
    alpha = 0.7, # making them a little transparent
    color = "white" # creates white border around each ridgeline 
  ) +
  # adding custom colors using scale_fill_manual
  scale_fill_manual(values = c(
    "Y" = "tomato2", # yes sick is red because this color gives sinus infection
    "N" = "skyblue3" # no not sick is blue because this color makes me feel calm, like i do when i do not have a sinus infection
  )) +
  # labeling title, subtitle, and axes
  labs(
    # title is general finding from the figure
    title = "Hydration patterns shifted ruring illness",
    # subtitle describes the variables being examined in this plot
    subtitle = "Distribution of daily water intake on sick and healthy days",
    # x axis is water intake in liters
    x = "Water intake (L)",
    # y axis is whether or not i have a sinus infection on the day the data is being collected
    y = "Sinus infection"
  ) +
  # setting a nice, classic theme that matches other plots with base size 14 for readability
  theme_classic(base_size = 14) +
  # removing legend, i do not find it to be necessary
  theme(legend.position = "none")

```

## **Bonus:** Visualization 4: Pie chart -- showing how many days I had a sinus infection for fun 

Note: I was honestly shocked at how having a sinus infection greatly affected the amount of water I drink compared to some of the other variables. I think emphasizing this in the infographic would be a fun and personal way to integrate this important variable into my statistical story. 

```{r}

sick_counts <- my_data_clean |>
  
  count(sick_y_n)

```

```{r}

# base layer ggplot using sick_counts data frame
ggplot(data = sick_counts,
       # using aes to set axes
       mapping = aes(x = "", # creates single stacked bar for pie chart
           y = n, # using column count for slice sizes in our pie
           fill = sick_y_n)) + # fill color based on whether or not i was sick
  # first layer geom_col which makes a bar chart --> pie chart
  geom_col(width = 1, # width 1 makes bars full width for pie chart
           color = "white") + # color white adds the white border between the pie slices as seen around the red slice
  # transforms bar chart into circle
  coord_polar(theta = "y") + # mapping y values into angles
  # custom coloring
  scale_fill_manual(values = c(
    # yes, sinus infection is red
    "Y" = "tomato2",
    # no, no sinus infection is blue
    "N" = "skyblue2"
  ),
    # setting legend labels 
    labels = c(
    # "Y" becomes "Yes" so readers can better understand legend
    "Y" = "Yes",
    # "N" becomes "No" so readers can better understand legend
    "N" = "No"
  )) +
  # using labs to set some titles
  labs(
    # title explains what the pie chart is about
    title = "How often illness appeared in daily life",
    # subtitle explains how this pie chart conveys proportion of days with and without sickness
    subtitle = "Proportion of days with and without a sinus infection",
    # colors filled by sickness, legend title is "sinus infection?" 
    fill = "Sinus infection?"
  ) +
  # theme void because it gets rid of basically all background elements, base size 14 for easy readability
  theme_void(base_size = 14)

```

